---
title: "Border Patrol Apprehensions:<br>
2010 vs. 2017"
author: "Han Nguyen, Carlos Echeverri, Nathan Mokhtarzadeh"
date: "February 19, 2018"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Introduction

#### Since Donald Trump became President, apprehensions at the US-Mexico border have been very low compared to previous years. 

* People have stopped illegally crossing the Southern border
    + due to the President's "rhetoric and aggressive push" to enforce immigration laws. 
* "Would-be" immigrants may be waiting and watching instead of crossing the border.


## In this project...

* We compare the number of border patrol apprehensions in 2010 and 2017 by months and sector to see if there has been a significant decrease in apprehensions
    + Big Bend
    + Del Rio
    + El Centro
    + El Paso
    + Laredo
    + Rio Grande Valley
    + San Diego
    + Tucson
    + Yuma


## Visualization of the Data

We compare the number of apprehensions in 2010 and 2017 by months and sector in a side by side barplot. The following slides show these barplots.

## Big Bend
```{r}
# read bp tables for each year
A2010 <- read.csv("bp appre 2010.csv", header = TRUE, stringsAsFactors = FALSE)
A2017 <- read.csv("bp appre 2017.csv", header = TRUE, stringsAsFactors = FALSE)


# get name of each sector
rownames(A2010) <- A2010[,1]
rownames(A2017) <- A2017[,1]


# function to get totals of each sector, month, and data as a whole
getTotals <- function(data) {
  data <-  subset(data, select= -c(Sector))
  rownames(data)
  
  data <- rbind(data, colSums(data))
  
  rownames(data)
  
  -length(rownames(data))  
  
  rownames(data) <- c(rownames(data)[-length(rownames(data))], "Total")
  
  data <- cbind(data,rowSums(data))
  
  colnames(data) <- c(colnames(data)[-length(colnames(data))], "Total")
  return(data)
  
}

# Append totals for data
A2010 <- getTotals(A2010)
A2017 <- getTotals(A2017)

# create side-by-side barplots comparing the 2010 and 2017 statistics by 
# month and sector
title <- rownames(A2010)[1]
    
both <- rbind(A2010[1,1:12], A2017[1,1:12])
    
row.names(both) <- c("2010", "2017")
    
barplot(as.matrix(both), beside = TRUE, col = c("red", "blue"), bty="n",xlab = 'Month',
        ylab = 'Number of Apprehensions', cex.names = 0.5 )
    
legend("topright", c("2010","2017"), pch=15,  col=c("red","blue"),  bty="n", 
        title = title)

```

## Del Rio
```{r}
title <- rownames(A2010)[2]
    
both <- rbind(A2010[2,1:12], A2017[2,1:12])
    
row.names(both) <- c("2010", "2017")
    
barplot(as.matrix(both), beside = TRUE, col = c("red", "blue"), bty="n",xlab = 'Month',
        ylab = 'Number of Apprehensions', cex.names = 0.5 )
    
legend("topright", c("2010","2017"), pch=15,  col=c("red","blue"),  bty="n", 
        title = title)
```

## El Centro
```{r}
title <- rownames(A2010)[3]
    
both <- rbind(A2010[3,1:12], A2017[3,1:12])
    
row.names(both) <- c("2010", "2017")
    
barplot(as.matrix(both), beside = TRUE, col = c("red", "blue"), bty="n",xlab = 'Month',
        ylab = 'Number of Apprehensions', cex.names = 0.5 )
    
legend("topright", c("2010","2017"), pch=15,  col=c("red","blue"),  bty="n", 
        title = title)
```

## El Paso
```{r}
title <- rownames(A2010)[4]
    
both <- rbind(A2010[4,1:12], A2017[4,1:12])
    
row.names(both) <- c("2010", "2017")
    
barplot(as.matrix(both), beside = TRUE, col = c("red", "blue"), bty="n",xlab = 'Month',
        ylab = 'Number of Apprehensions', cex.names = 0.5 )
    
legend("topright", c("2010","2017"), pch=15,  col=c("red","blue"),  bty="n", 
        title = title)
```

## Laredo
```{r}
title <- rownames(A2010)[5]
    
both <- rbind(A2010[5,1:12], A2017[5,1:12])
    
row.names(both) <- c("2010", "2017")
    
barplot(as.matrix(both), beside = TRUE, col = c("red", "blue"), bty="n",xlab = 'Month',
        ylab = 'Number of Apprehensions', cex.names = 0.5 )
    
legend("topright", c("2010","2017"), pch=15,  col=c("red","blue"),  bty="n", 
        title = title)
```

## Rio Grande Valley
```{r}
title <- rownames(A2010)[6]
    
both <- rbind(A2010[6,1:12], A2017[6,1:12])
    
row.names(both) <- c("2010", "2017")
    
barplot(as.matrix(both), beside = TRUE, col = c("red", "blue"), bty="n",xlab = 'Month',
        ylab = 'Number of Apprehensions', cex.names = 0.5 )
    
legend("topright", c("2010","2017"), pch=15,  col=c("red","blue"),  bty="n", 
        title = title)
```

## San Diego
```{r}
title <- rownames(A2010)[7]
    
both <- rbind(A2010[7,1:12], A2017[7,1:12])
    
row.names(both) <- c("2010", "2017")
    
barplot(as.matrix(both), beside = TRUE, col = c("red", "blue"), bty="n",xlab = 'Month',
        ylab = 'Number of Apprehensions', cex.names = 0.5 )
    
legend("topright", c("2010","2017"), pch=15,  col=c("red","blue"),  bty="n", 
        title = title)
```

## Tucson
```{r}
title <- rownames(A2010)[8]
    
both <- rbind(A2010[8,1:12], A2017[8,1:12])
    
row.names(both) <- c("2010", "2017")
    
barplot(as.matrix(both), beside = TRUE, col = c("red", "blue"), bty="n",xlab = 'Month',
        ylab = 'Number of Apprehensions', cex.names = 0.5 )
    
legend("topright", c("2010","2017"), pch=15,  col=c("red","blue"),  bty="n", 
        title = title)
```

## Yuma
```{r}
title <- rownames(A2010)[9]
    
both <- rbind(A2010[9,1:12], A2017[9,1:12])
    
row.names(both) <- c("2010", "2017")
    
barplot(as.matrix(both), beside = TRUE, col = c("red", "blue"), bty="n",xlab = 'Month',
        ylab = 'Number of Apprehensions', cex.names = 0.5 )
    
legend("topright", c("2010","2017"), pch=15,  col=c("red","blue"),  bty="n", 
        title = title)
```

## Overview of Barplots

* Much higher number of apprehensions in 2010 between the months of February and May
* Higher number of apprehensions in 2017 between September and December
* Tucson and San Diego had the most change
* Overall large decrease in number of apprehensions from 2010 to 2017

## Statistical Testing
### Most Apprehensions by Sector
In 2010, Tucson had the most apprehensions, and in 2017, Rio Grande Valley had the most apprehensions.
```{r}
customTest <- function(earlyData, newData) {
  #Assumes getTotals function has been run on data
  #Data for old data input
  maximumOld <- max(earlyData[1:9,13])
  maxRowIndexOld <- which.max(earlyData[1:9,13])
  earlyDataMean <- rowMeans(earlyData[maxRowIndexOld,1:12])

  
  #Data for new data input
  maximumNew <- max(newData[1:9,13])
  maxRowIndexNew <- which.max(newData[1:9,13])
  newDataMean <- rowMeans(newData[maxRowIndexNew,1:12])

  
  t.test(earlyData[maxRowIndexOld,1:12], newData[maxRowIndexNew,1:12])
  
}
customTest(A2010,A2017)
```

----

* The t-test gave a t-score of 2, and a p-value of 0.06. Using an alpha of 0.05, we fail to reject the null hypothesis since the p-value is greater than alpha.
* The difference in the means of apprehensions in 2010 and 2017 are not significantly different
* The confidence interval, which is (-380, 12820), includes 0, meaning there is not a significant difference in apprehensions

----

### Most Apprehensions in 3 Month Period
In 2010, March, April, and May had the most apprehensions, and in 2017, October, November, and December had the most apprehensions.
```{r}
#Get top three months
topMonths <- function(data){
  copy <- data
  copy <- sort(copy[10,1:12], decreasing = TRUE)[1:3]
  copy
}

# T test for top three months in apprehensions
compareThreeMonthPeriod <- function(earlyData, newData) {
  earlyDataThree <- topMonths(earlyData)
  newDataThree <- topMonths(newData)
  # get indices of these months
  earlyColName1 <- grep(colnames(earlyDataThree[1]), colnames(earlyData))
  earlyColName2 <- grep(colnames(earlyDataThree[2]), colnames(earlyData))
  earlyColName3 <- grep(colnames(earlyDataThree[3]), colnames(earlyData))
  earlyIndex <- min(earlyColName1,earlyColName2,earlyColName3)
  
  newColName1 <- grep(colnames(newDataThree[1]), colnames(newData))
  newColName2 <- grep(colnames(newDataThree[2]), colnames(newData))
  newColName3 <- grep(colnames(newDataThree[3]), colnames(newData))
  newIndex <- min(newColName1,newColName2,newColName3)

  
  earlyMean <- mean(c(earlyData[1:9,earlyIndex],earlyData[1:9,earlyIndex+1],
                      earlyData[1:9,earlyIndex+2]))
  newMean <- mean(c(newData[1:9,newIndex],newData[1:9,newIndex+1],newData[1:9,newIndex+2]))
  
  
  
  t.test(c(earlyData[1:9,earlyIndex],earlyData[1:9,earlyIndex+1],
           earlyData[1:9,earlyIndex+2]),c(newData[1:9,newIndex],newData[1:9,newIndex+1],
                                          newData[1:9,newIndex+2]))
  
  
}

compareThreeMonthPeriod(A2010,A2017)
```

----

* The t score was 0.5 and the p-value was 0.6, which is way above the alpha of 0.05. Since the p-value was above alpha, we fail to reject the null hypothesis
* The true difference in means is equal to 0, which means there is no significant difference in the means comparing most apprehensions in 3 month periods
* The confidence interval says the same since it is (-3120,5120), which includes 0, meaning the differences in means are insignificant

----

### Most Apprehensions Overall
Because we believed there still may be a significant difference, we took a t-test of each data as a whole in 2010 and 2017.
```{r}
t.test(A2010, A2017)
```

----

* Surprisingly, there was not a significant difference in the means of apprehensions
* The t test gave a t score of 0.9 and the p-value was 0.3, which is above the alpha of 0.05, meaning we fail to reject the null hypothesis, so there is no true difference between apprehensions

## Time Series
```{r}
# time series chart for monthly summaries
ts <- readClipboard(raw = FALSE)

ts1 <- read.table("clipboard", sep = "\t")

ts2 <- rev(as.vector(t(ts1)))

ts3 <- ts(ts2, start = c(2000,10), frequency=12)

ts.plot(ts3, gpars=list(xlab="year", ylab="Apprehensions", lty=c(1:3)), col = 'blue')
```

----

* The time series chart shows a decrease in apprehensions throughout the years from 2000 to 2017 
* There seems to be a large decrease, although our statistical tests say otherwise for 2010 and 2017

## Conclusion
* There is a decrease in apprehensions throughout the years
    + because less "would-be" immigrants are crossing the border since the president has          been changing immigration laws
* The t-tests don't show a significant difference between mean of apprehensions in 2010 and 2017
    + this may be because we are only comparing the maximum number of apprehensions by            sectors or months in each year, disregarding what month and sector they actually are      
    + if there was a comparison between each sector and each month, there may be a bigger         significant difference.
